Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway

### 2023-07-27 analysis/paper/paper.Rmd
Discussion

The potentially immense value of insights into population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates-as-data methodology [@rick1987] has undoubtedly contributed to the popularity of the approach. Several limitations and forms of criticism have, however, been directed at these procedures. Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a direct connection between the frequency of ^14^C-dates and population dynamics [e.g. @carleton2021; @torfing2015; @ward2021]. What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics and other variables that might impact these [@bevan2017; @french2021; @palmisano2017; @palmisano2021]. In this paper, we aim to report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD). We discuss how the data compare with each other as well as how it can be used to understand cultural historical development in the region through the Mesolithic, c. 9500--4000 BCE.

# Background
Following the melting of the Fennoscandian Ice Sheet, large parts of the post-glacial landscape of Northern Scandinavia has been characterised by dramatic isostatic uplift that has led to a net sea-level fall throughout the Holocene, despite eustatic sea-level rise [e.g. @mörner1979]. The majority of Mesolithic sites along the Norwegian Skagerak coast are interpreted as having been shore bound or at least located close to the contemporaneous shoreline at the time of use. As coastal foragers in this region appear to have predominantly settled on or close to the contemporaneous shoreline during early and mid-Holocene, this can be utilised to assign an approximate date to the sites. This is done by coupling the present-day altitude of the sites with reconstructions of past shoreline displacement -- a method known as shoreline dating. method has a long tradition in Norwegian archaeology [@brogger1905], and is especially useful for dating the large number of surveyed archaeological sites when temporal data that often follow with an excavation, such as radiocarbon dates or typological indicators in artefact inventories, are not available.  Previous studies shave suggested that shoreline dated sites can be used as a proxy to study temporal variation in population and settlement patterns [@solheim2018; @tallavaara2020; @jorgensen2020].

The frequency of shoreline dated sites has been compared to RSPDs in the past by finding point estimates of shoreline dates that are aggregated in somewhat arbitrary bins of 200 or 500 years. This therefore does not consider the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method. In a recent study, @roalkvam2023 has presented a probabilistic method for shoreline dating that takes these parameters into account, thus setting the stage for a more refined investigation of the relationship between frequency of ^14^C-dates and shoreline dated sites. The parametrisation of the method was based on simulating the distance between sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the northeast to Arendal municipality in the southwest (see map in Figure \@ref(fig:map)). The results of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline [see also @nielsen2021a], followed by a clear break around 2500 BCE, at which point shoreline dating appears to lose its utility. Thus, this geographical and temporal limit are also used for this study.   
The present paper is decidedly exploratory. While some thoughts concerning the relationship between the variables outlined above underlie the analysis, these could not be instantiated as concrete hypotheses. Rather, we explore the data by using a set of standard models to begin to unpack patterns and grasp the relationship between the variables. Furthermore, the dates-as-data approach is dependent on there being a link between how prehistoric people generated material that become ^14^C samples, and reflect population dynamics [cf- @freeman2018]. The sum of shoreline dated sites, on the other hand, is determined by site frequency, and its use as a population proxy is therefore dependent on there being a connection between site count and population size and density. If a comparison of these proxies do not, or only partially correspond, it is thus an open question what factors have impacted either distribution to cause this discrepancy, and which measure, if any, reflects true population dynamics. The issue will therefore initially demand an open and exploratory approach where a multitude of explanatory and confounding effects can be drawn on to suggest explanations of any observed pattern.

@@

To what degree the radiocarbon record is determined by past population numbers might vary geographically and chronologically, based on variation in investigatory and taphonomic factors [@surovell2009b; @bluhm2019], as well as cultural processes within prehistoric populations. One example of the latter is the difference that might exist between farmer and forager populations, where @freeman2018 have suggested that an increased per capita energy consumption introduced with farming means that ^14^C-dates should not be weighted equally when making relative population estimates across such populations. Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns, and settlement nucleation and dispersion [@palmisano2017], as well as variation in archaeological investigations and what has been taken to consitute a 'site'. However, these could be considered theoretical issues that have implications for how fluctuations in these proxies should be interpreted, and possibly weighted differently according some criteria. Before any such fluctuations are given any substantive interpretation, however, there is a host of methodological issues that have to be considered [see @carleton2020; @crema2022 for a review].

The most critical of these follow from the fact that the summation of the probabilities associated with the dates for the SPDs is not a statistically coherent procedure. This is because the summed probabilities can no longer been seen as probabilities, but rather represent the combination of events and chronological uncertainties, making the two indistinguishable, and rendering the interpretation of the resulting sum difficult [@blackwell2003; @crema2022]. As @timpson2021[2] put it: 'the SPD is not *the* single best explanation of the data, nor even *a* single explanation of the data, but rather a conflation of many possible explanations simultaneously, each of which is mired by the artefacts inherited from the calibration wiggles.' The SPD is not a model. It is the combined representation of a range of possible explanations for the data -- the frequency of dated events combined with the variable uncertainty associated with these [@carleton2021]. This means that a direct visual inspection of SPDs can be misleading, that they cannot be directly analysed to draw inferences on population dynamics, nor can they be directly compared to other time-series data [@carleton2018; @timpson2021; @crema2022]. While this problem can never be entirely resolved, a range of approaches have been developed in an attempt to work around this issue.

@@

## Model comparison and continuous piece-wise linear models

The procedure outlined above represent the standard approach in efforts to model population dynamics based on RSPDs. This can be helpful for determining whether the population patterns has followed the trajectories of standard theoretical models. It is also a useful first step here also for the treatment of shoreline dated sites, as few assumptions concerning the overall development of this proxy could be made. However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and, as mentioned above, deviations from these rejected null-models do not in themselves allow for any interpretations of the local deviations, nor a subsequent direct analysis of the SPDs themselves. As a result, and by improving on @goldstein2016, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics by use of RSPDs, published with the accompanying R package  *ADMUR*. This involves fitting continuous piece-wise linear (CPL) models to the data, in addition to the standard models, subjecting these to a procedure of model comparison, and finally evaluating the best model using a goodness-of-fit test, in line with the Monte Carlo approach outlined above. The CPL models are defined by discrete linear periods of development between hinge points that mark points in time in which the trajectory of the population proxy changes direction. This allows for the identification of key demographic events and a conservative description of population dynamics between these. The approach thus provides a clearer inferential justification for a further analysis and substantive interpretation of the identified best model. Following the evaluation of standard models as outlined above, the data were subsequently subjected to the approach presented by @timpson2021.

@@

Sites surveyed by means of test-pitting between Horten and Arendal were initially retrieved from the national heritage database Askeladden [@directorate2018], totalling at `r nrow(surveyed)` records. The records were then manually reviewed and given a quality score based on the criteria in Table \@ref(tab:tab1), indicating the degree to which the spatial location and extent of the sites is believed to be represented in the geometries available in the database [see also @roalkvam2020]. All sites with a quality score of 4 or worse were excluded from further analysis. Any sites situated at elevations that result in a shoreline date earlier than 9469 BCE were then excluded. This marks the latest start-date among the employed displacement curves, and no sites are yet verifiably older than around 9300 BCE [@glorstad2016]. Data on excavated sites was originally compiled for @roalkvam2023 and has been compared with site data as listed in @damlien2021 and @nielsen2019. Only excavated sites with available spatial data in Askeladden or local databases at the Museum of Cultural History of the University of Oslo were included in the analysis. The 102 excavated sites in the dataset without relevant ^14^C-dates and that were originally shoreline dated in the reports were included in the SSPD along with the retained surveyed sites. This gave a total of  shoreline dated sites in the final SSPD.

The borders of the municipalities within which the shoreline dated sites are located were used to limit the radiocarbon sample. Radiocarbon dates were taken from @roalkvam2023 and @solheim. Dates done on food crusts were then excluded due the issue of marine reservoir effects [@nielsen2019, 83], resulting in a final total of ^14^C-dates. Following both from the point made by @freeman2018 concerning the comparison of radiocarbon dates from populations with different economic modes, as noted above, and from the fact that shoreline dating appears less reliable after 4000 BCE and appears to loose its utility c. 2500 BCE, the Mesolithic is the main focus of this analysis. Consequently, probabilities falling later than 2500 BCE were excluded from the analysis. Probabilities falling after 4000 BCE and the first, possible small-scale introduction of agriculture in south-eastern Norway were retained to account for edge-effects, but these results should be treated with care for the reasons noted above.

@@

Four geological reconstructions of shoreline displacement in the region lays the foundation for the method as implemented here (Figure \@ref(fig:shoredate)B). These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory for relative sea-level change has been the same [see @svendsen1987]. The first step in the dating procedure is to interpolate the shoreline displacement to the site to be dated. This is done by inverse distance weighting [e.g. @conolly2020], interpolating the relative sea-level change to the location of an archaeological site, based on its distance from the isobases of the displacement curves.

@@

## Continuous piece-wise linear models
The procedures outlined above represent standard approaches when modelling population dynamics in archaeology. These can be helpful for determining whether the population behaviour has followed the trajectories of common theoretical population models, and is an especially useful exercise here as few assumptions concerning the behaviour and overall development of number of shore-bound sites over time could be made. However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and deviations from these rejected null-models do not in themselves allow for any interpretations of the deviations, nor a subsequent direct analysis of the SPDs -- although this has frequently been the case. As a result, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics. This involves fitting continuous piece-wise linear (CPL) models in addition to the standard models to the data, subjecting these to a procedure for model comparison, and finally testing the best model using the approach outlined above. The benefit of the CPL models is that these define discrete periods between hinge points that mark points in time in which the trajectory of relative population size changes direction, thus identifying key events and characteristics of the population dynamics between these. Following the evaluation of standard models outlined above, the data was therefore subsequently subjected to the approach outlined by @timpson2021.

@@

For each dating method, the relative performance of the exponential, logistic, uniform and CPL models was compared. This was done by finding the overall relative log-likelihoods of the models using the *loglik* function from *ADMUR*. For any given model, this involves finding the likelihood for each shoreline date against the null-model, and in the case of the ^14^C-dates, each site-phase. The overall model likelihood is the product of each of these individual likelihoods and is not, critically [@timpson2021, 3], found by estimating the likelihood using the final SPD. From these likelihoods, the Bayesian (or Schwarz) information criterion (BIC) was then found for each model. This penalises the models for the number of parameters in use, to avoid over-fitting, and can be used to compare the relative performance of the models.

@@

```{r spdmc, echo = FALSE, fig.cap = "Monte Carlo simulation for shoreline and radiocarbon dates. A) Summed probability of shoreline dated sites (n = 921) compared to an exponential null model. B) SSPD compared to a logistic null model. C) SSPD compared to a uniform null model. D) Summed probability of calibrated radiocarbon dates (n = 310, bins = 134) compared to an exponential null model. E) RSPD compared to a logistic null model. F) RSPD compared to a uniform null model.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/mc.png"))
```
For the RSPD, the logistic model achieved a p-value of 0.108, and could therefore not be rejected. Based on the available data, this model therefore represents the best explanation of the data. Given the failure to reject the logistic model, the subsequent step of also including a comparison with CPL models was not strictly necessary, but the result of this procedure further substantiates the finding as the logistic model is the best among the considered candidates.

For the SSPD all of the standard models could be rejected. The result of subjecting the SSPD to the ADMUR workflow identified 4-CPL as the best model. An important caveat here is that models with a complexity beyond 4-CPL consistently failed to converge. A fuller reporting of this convergence issue, including the exploration of different parameter settings, different sample sizes and alternative optimization algorithms is given in the supplementary material. Subjecting the 4-CPL model to the goodness-of-fit test gave p =, meaning that the data nonetheless could be expected under this model. The accepted models that underlies the comparison to follow are therefore the logistic model for the RSDP and the 4-CPL for the SSPD.

<!-- We can also add that accepting this model means that interpreting any deviations from the Monte Carlo enevelope is inappropriate, although this has sometimes been the case. -->
A notable pattern in Figure \@ref(fig:spd_mc) is the lack of data in the RSPD before c. 8500--8300 BCE while the SSPD has a marked peak around 9000 BCE. Comparing the the directions of the standard models in the figure, the logistic model fit to the RSPD in Figure \@ref(fig:spd_mc) indicates an overall uniform distribution in the frequency of radiocarbon dates following the initial jump just before 8000 BCE. This is contrasted to a general decrease in shoreline dated sites after the initial rapid increase from around 9500 BCE. Although they could be rejected as explaining the data, this is indicated by the decaying exponential function (Figure \@ref(fig:spd_mc)A) and the initial under-prediction followed by regions of over-prediction with the logistic and uniform models (Figure \@ref(fig:spd_mc)B and C).

It is worth noting that there appears to be a corresponding peak in both proxies just after 8000 BCE and similarly a drop just before 6000 BCE. However, given that the logistic model cannot be rejected for the RSPD, it is inappropriate to interpret any deviations from the Monte Carlo envelope, and no analytical weight can be placed on these deviations -- at least given the present data and modelling efforts.

@@

It is established that Stone Age sites in the region date to the period from before the earliest ^14^C-dates, and at least from as far back as around 9300 BCE [e.g. @damlien2018b; @glorstad2016]. While the SSPD starts from c. 9500 BCE, the RSPD show no evidence of human activity until c. 8500–8300 BCE. What has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations through the years. These range from from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs for fuel or use of rapidly deteriorating seal blubber for burning, a preference for consuming raw, dried or fermented food, and possibly a lack of reliable access to firewood [e.g. @damm2022]. While taphonomic and ecological factors cannot be written off, we view it as likely that the initial jump at the start of RSPD, combined with the magnitude of the SSPD in the period preceding this, suggest that cultural rather than environmental factors are central drivers behind the observed pattern.

The considered models indicate that after the initial jumps, some process of decrease in shoreline dated site frequency occurs throughout the period, contrasted by a general stationary frequency of ^14^C-dated material. An immediate explanation of this mismatch could be that the SSPD mainly reflects variations in land-use and mobility patterns, while the RSPD mainly reflects population numbers. A common interpretation of the Early Mesolithic in the region is that the population was small but highly mobile, reflected in large numbers of short lived sites along the coast [e.g. @bjerck2017]. This is commonly seen in relation to homogeneous lithic technology over large parts of Scandinavia in this period, and what is believed to have been a rapid colonisation of the Norwegian coastline. High mobility has been seen as necessary both to maintain demographic viability among colonising groups and for it to be possible to maintain this technological uniformity over large areas [@rowley-conwy2017]. The initial peak in the SSPD could thus be congruent with a process of rapid colonisation of the Norwegian coast, coupled with a high degree of mobility, leading to the overall high site-count in the earliest period. At the same time, the lacking signal in the RSPD for this earliest period could be taken as evidence for low population numbers during the first settlement period following the deglaciation.

@berg-hansen2022 have argued that there is an increase in sites between c. 8000--7000 BCE in the northern part of the study area of this paper. However, their analysis covers a limited area and is based on the aggregation of point-based estimates of shoreline dates, which are aggregated within periods of variable length. Apart from the limitations with using point-based estimates for shoreline dates, the use of disjoint time-intervals of variable duration could cause issues for comparing the relative counts across these units [@bevan2021].

In contrast to the findings of @berg-hansen2022, the analysis undertaken here instead indicates a drop in the number of sites through this time-interval. It has previously been suggested that a change in mobility patterns occurs from some time around 8300 BCE, involving a reduction in residential mobility [@roalkvam2022; @damlien2018b]. Furthermore, both genetic data and technological analyses of lithic inventories indicates that the initial phase of human colonisation from the south is followed by an influx, and possible mixing, of people from the east at the transition to the Middle Mesolithic period around the same time [@gunther2018; @manninen2021]. If this narrative is coupled with the data observed here, these migratory events would appear to result in a relatively sudden drop in site-count as indicated in the SSPD, which in turn could be related to a change in mobility patterns. The end of this drop corresponds to the first ^14^C-dates in the RSPD around 8300 BCE -- giving further credence to the suggestion that the appearance of material to be ^14^C-dated is mainly related to cultural and demographic processes rather than taphonomic factors.

The development after 4000 BCE should be treated with further care, following from questions concerning the reliability of shoreline dating after 4000 BCE [cf. @roalkvam2023].

>>>>

The potentially immense value of insights into population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates-as-data methodology [@rick1987] has undoubtedly contributed to the popularity of the approach. Several limitations and forms of criticism have, however, been directed at these procedures. Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a direct connection between the frequency of ^14^C-dates and population dynamics [e.g. @carleton2021; @torfing2015; @ward2021]. What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics and other variables that might impact these [@bevan2017; @french2021; @palmisano2017; @palmisano2021]. In this paper, we aim to report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD). We discuss how the data compare with each other as well as how it can potentially be used to understand cultural historical development in the region through the Mesolithic, c. 9500--4000 BCE.

# Background
Following the melting of the Fennoscandian Ice Sheet, large parts of the post-glacial landscape of Northern Scandinavia has been characterised by dramatic isostatic uplift that has led to a net sea-level fall throughout the Holocene, despite eustatic sea-level rise [e.g. @mörner1979; @steffen2011]. The majority of Mesolithic sites along the Norwegian Skagerak coast are interpreted as having been shore bound or at least located close to the contemporaneous shoreline at the time of use. As coastal foragers in this region appear to have predominantly settled on or close to the contemporaneous shoreline during early and mid-Holocene, this can be utilised to assign an approximate date to the sites. This is done by coupling the present-day altitude of the sites with reconstructions of past shoreline displacement -- a method known as shoreline dating. Shoreline dating has a long tradition of use in Norwegian archaeology [@brogger1905], and is especially useful for dating the large number of surveyed archaeological sites where temporal data that often follow with an excavation, such as radiocarbon dates or typological indicators in artefact inventories, are not available. Previous studies shave suggested that shoreline dated sites can be used as a proxy to study temporal variation in population and settlement patterns [@solheim2018; @tallavaara2020; @jorgensen2020].

Previously, the frequency of shoreline dated sites has been compared to RSPDs by finding point estimates of shoreline dates that are aggregated in somewhat arbitrary bins of 200 or 500 years. This therefore does not consider the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method. In a recent study, @roalkvam2023 has presented a probabilistic method for shoreline dating that takes these parameters into account, thus setting the stage for a more refined investigation of the relationship between frequency of ^14^C-dates and shoreline dated sites. The parametrisation of the method was based on simulating the distance between 66 sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the northeast to Arendal municipality in the southwest (see map in Figure \@ref(fig:map)). The results of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline [see also @nielsen2021a], followed by a clear break around 2500 BCE, at which point shoreline dating appears to lose its utility. Thus, this geographical and temporal limit are also used for this study.   
The present paper is decidedly exploratory. While some thoughts concerning the relationship between the variables outlined above underlie the analysis, these could not be instantiated as concrete hypotheses. Rather, we explore the data by using a set of standard models to begin to unpack patterns and grasp the relationship between the variables. The dates-as-data approach is dependent on there being a link between how prehistoric people generated material that become ^14^C samples and population dynamics [cf. @freeman2018]. The sum of shoreline dated sites, on the other hand, is determined by site frequency, and its use as a population proxy is therefore dependent on there being a connection between site count and population density. If a comparison of these proxies do not, or only partially correspond, it is thus an open question what factors have impacted either distribution to cause this discrepancy, and which measure, if any, reflects true population dynamics. The issue will therefore initially demand an open and exploratory approach where a multitude of explanatory and confounding effects can be drawn on to suggest explanations of any observed pattern.

@@

To what degree the radiocarbon record is determined by past population numbers might vary geographically and chronologically, based on variation in investigatory and taphonomic factors [@surovell2009b; @bluhm2019], as well as cultural processes within prehistoric populations. One example of the latter is the difference that might exist between farmer and forager populations, where @freeman2018 have suggested that an increased per capita energy consumption introduced with farming means that ^14^C-dates should not be weighted equally when making relative population estimates across such populations. Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns, and settlement nucleation and dispersion [@palmisano2017], as well as variation in archaeological investigations and what has been taken to constitute a 'site'. However, these could be considered theoretical issues that have implications for how fluctuations in these proxies should be interpreted, and possibly weighted differently according some criteria. Before any such fluctuations are given any substantive interpretation, however, there is a host of methodological issues that have to be considered [see @carleton2020; @crema2022 for a review].

The most critical of these follow from the fact that the summation of the probabilities associated with the dates for SPDs is not a statistically coherent procedure. This is because the summed probabilities can no longer been seen as probabilities, but rather represent the combination of events and chronological uncertainties, making the two indistinguishable, and rendering the interpretation of the resulting sum difficult [@blackwell2003; @crema2022]. As @timpson2021[2] put it: 'the SPD is not *the* single best explanation of the data, nor even *a* single explanation of the data, but rather a conflation of many possible explanations simultaneously, each of which is mired by the artefacts inherited from the calibration wiggles.' The SPD is not a model. It is the combined representation of a range of possible explanations for the data -- the frequency of dated events combined with the variable uncertainty associated with these [@carleton2021]. This means that a direct visual inspection of SPDs can be misleading, that they cannot be directly analysed to draw inferences on population dynamics, nor can they be directly compared to other time-series data [@carleton2018; @timpson2021; @crema2022]. While this problem can never be entirely resolved, a range of approaches have been developed in an attempt to work around this issue.

@@

The procedure outlined above represent the standard approach in efforts to model population dynamics based on RSPDs. This can be helpful for getting an impression of the overall developments of population proxies, and is an especially useful first step in the treatment of shoreline dated sites done here, as few assumptions concerning this overall development could be made. However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature and it is worth underscoring, as mentioned above, that deviations from these rejected null-models do not in themselves allow for any interpretations of the local deviations, nor a subsequent direct analysis of the SPDs themselves.

<!-- ## Model comparison and continuous piece-wise linear models -->

<!-- The procedure outlined above represent the standard approach in efforts to model population dynamics based on RSPDs. This can be helpful for determining whether the population patterns has followed the trajectories of standard theoretical models. It is also a useful first step here also for the treatment of shoreline dated sites, as few assumptions concerning the overall development of this proxy could be made. However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and, as mentioned above, deviations from these rejected null-models do not in themselves allow for any interpretations of the local deviations, nor a subsequent direct analysis of the SPDs themselves. As a result, and by improving on @goldstein2016, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics by use of RSPDs, published with the accompanying R package  *ADMUR*. This involves fitting continuous piece-wise linear (CPL) models to the data, in addition to the standard models, subjecting these to a procedure of model comparison, and finally evaluating the best model using a goodness-of-fit test, in line with the Monte Carlo approach outlined above. The CPL models are defined by discrete linear periods of development between hinge points that mark points in time in which the trajectory of the population proxy changes direction. This allows for the identification of key demographic events and a conservative description of population dynamics between these. The approach thus provides a clearer inferential justification for a further analysis and substantive interpretation of the identified best model. Following the evaluation of standard models as outlined above, the data were subsequently subjected to the approach presented by @timpson2021. -->

@@

Sites surveyed by means of test-pitting between Horten and Arendal were initially retrieved from the national heritage database Askeladden [@directorate2018], totalling at `r nrow(surveyed)` records. The records were then manually reviewed and given a quality score based on the criteria in Table \@ref(tab:tab1), indicating the degree to which the spatial location and extent of the sites is believed to be represented in the geometries available in the database [see also @roalkvam2020]. All sites with a quality score of 4 or worse were excluded from further analysis. Any sites situated at elevations that result in a shoreline date earlier than 9465 BCE were then excluded. This marks the latest start-date among the employed displacement curves, and no sites are yet verifiably older than around 9300 BCE [@glorstad2016]. Data on excavated sites was originally compiled for @roalkvam2023 and has been compared with site data as listed in @damlien2021 and @nielsen2019. Only excavated sites with available spatial data in Askeladden or local databases at the Museum of Cultural History of the University of Oslo were included in the analysis. The 102 excavated sites in the dataset without relevant ^14^C-dates that were originally shoreline dated in the excavation reports were included in the SSPD along with the retained surveyed sites. This gave a total of 921 shoreline dated sites in the final SSPD.

The borders of the municipalities within which the shoreline dated sites are located were used to limit the radiocarbon sample. Radiocarbon dates were taken from @roalkvam2023 and @solheim. Dates done on food crusts were then excluded due the issue of marine reservoir effects [@nielsen2019, 83], resulting in a total of ^14^C-dates. Following both from the point made by @freeman2018 concerning the comparison of radiocarbon dates from populations with different economic modes, as noted above, and from the fact that shoreline dating appears less reliable after 4000 BCE and appears to loose its utility c. 2500 BCE, the Mesolithic is the main focus of this analysis. Consequently, probabilities falling later than 2500 BCE were excluded from the analysis, further reducing the sample of ^14^C-dates to. Probabilities falling after 4000 BCE and the first possible small-scale introduction of agriculture in south-eastern Norway were retained to account for edge-effects, but these results should be treated with care for the reasons noted above.

@@

Four geological reconstructions of shoreline displacement in the region lays the foundation for the method as implemented here (Figure \@ref(fig:shoredate)B). These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory for relative sea-level change has been the same [see @svendsen1987]. The first step in the dating procedure is to interpolate the shoreline displacement to the site to be dated. This is done by inverse distance weighting [e.g. @conolly2020], where the relative sea-level change is interpolated to the location of an archaeological site based on its distance from the isobases of the geological displacement curves.

@@

<!-- ## Continuous piece-wise linear models -->
<!-- The procedures outlined above represent standard approaches when modelling population dynamics in archaeology. These can be helpful for determining whether the population behaviour has followed the trajectories of common theoretical population models, and is an especially useful exercise here as few assumptions concerning the behaviour and overall development of number of shore-bound sites over time could be made. However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature, and deviations from these rejected null-models do not in themselves allow for any interpretations of the deviations, nor a subsequent direct analysis of the SPDs -- although this has frequently been the case. As a result, @timpson2021 have recently suggested an alternative framework for modelling past population dynamics. This involves fitting continuous piece-wise linear (CPL) models in addition to the standard models to the data, subjecting these to a procedure for model comparison, and finally testing the best model using the approach outlined above. The benefit of the CPL models is that these define discrete periods between hinge points that mark points in time in which the trajectory of relative population size changes direction, thus identifying key events and characteristics of the population dynamics between these. Following the evaluation of standard models outlined above, the data was therefore subsequently subjected to the approach outlined by @timpson2021. -->

@@

For each dating method, the relative performance of the exponential, logistic and uniform models was compared. This was done by finding the overall relative log-likelihoods of the models using the *loglik* function from *ADMUR*. For any given model, this involves finding the likelihood for each shoreline date against the null-model, and in the case of the ^14^C-dates, each site-phase. The overall model likelihood is the product of each of these individual likelihoods and is not, critically [@timpson2021, 3], found by estimating the likelihood using the final SPD. From these likelihoods, the Bayesian (or Schwarz) information criterion (BIC) was then found for each model. This penalises the models for the number of parameters in use, to avoid over-fitting, and can be used to compare the relative performance of the models.

@@

```{r spdmc, echo = FALSE, fig.cap = "Monte Carlo simulation for shoreline and radiocarbon dates. A) Summed probability of shoreline dated sites (n = 921) compared to an exponential null model. B) SSPD compared to a logistic null model. C) SSPD compared to a uniform null model. D) Summed probability of calibrated radiocarbon dates (n = 310, bins = 134) compared to an exponential null model. E) RSPD compared to a logistic null model. F) RSPD compared to a uniform null model.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/mc.png"))
```
For the RSPD, the logistic model achieved a p-value of 0.072, and could therefore not be rejected. As is also indicated by the BIC values, this model represents the best explanation of the data, while the failure to reject the model means that the data could be expected under this logistic development. This is characterised by a rapid growth from the start of the RSPD which plateaus at around 7500 BCE and then remains stable for the remainder of the period.

<!-- Given the failure to reject the logistic model, the subsequent step of also including a comparison with CPL models was not strictly necessary, but the result of this procedure further substantiates the finding as the logistic model is the best among the considered candidates. -->

All of the standard models could be rejected for the SSPD. Of the contrasted models, the BIC values indicate that the exponential model is the best candidate. The negative direction indicates that some overall process of reduction in the frequency of shoreline dated sites occurs over time, while the rejection of the model and clear deviations from the exponential patterns indicates that this model does not account for the entire development. Some significant deviations occurs with the model over-predicting the frequency of sites in the period from c. X--9000 BCE, followed by a significant positive peak just after 9000 BCE. The process of decline is then interrupted by positive deviations at, and with a smaller deviation around 7000 BCE. This is closely followed by a negative deviation of more magnitude than that expected under the exponential model between X and X BCE. Finally, a positive deviation occurs again at X--5000 BCE. While all of these deviations should be viewed with caution, the small magnitude and questions concerning the reliability of shoreline dating after 4000 BCE [cf. @roalkvam2023] means that the negative deviations that occur around 3000 BCE are disregarded here.

<!-- For the SSPD all of the standard models could be rejected. The result of subjecting the SSPD to the ADMUR workflow identified 4-CPL as the best model. An important caveat here is that models with a complexity beyond 4-CPL consistently failed to converge. A fuller reporting of this convergence issue, including the exploration of different parameter settings, different sample sizes and alternative optimization algorithms is given in the supplementary material. Subjecting the 4-CPL model to the goodness-of-fit test gave p =, meaning that the data nonetheless could be expected under this model. The accepted models that underlies the comparison to follow are therefore the logistic model for the RSDP and the 4-CPL for the SSPD.      -->

<!-- We can also add that accepting this model means that interpreting any deviations from the Monte Carlo enevelope is inappropriate, although this has sometimes been the case. -->

@@

It is established that Stone Age sites in the region date to the period from before the earliest ^14^C-dates, and at least from as far back as around 9300 BCE [e.g. @damlien2018b; @glorstad2016]. While the SSPD achieved here starts from c. 9500 BCE, the RSPD shows no evidence of human activity until c. 8500 BCE. What has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations through the years. These range from from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs or rapidly deteriorating seal blubber as fuel for burning, a preference for consuming raw, dried or fermented food, and possibly a lack of reliable access to firewood [see @damm2022]. While taphonomic and ecological factors cannot be excluded as part of the picture, we view it as likely that the rapid growth at the start of RSPD and the magnitude of the SSPD in the period preceding this suggest that cultural and demographic as opposed to taphonomic and environmental factors are central drivers behind the observed pattern.

The considered models indicate that after the initial jumps of each proxy, some process of decrease in shoreline dated site frequency occurs throughout the period, contrasted by a general stationary frequency of ^14^C-dated material. An immediate explanation of this mismatch could be that the SSPD is more heavily impacted by variation in land-use and mobility patterns, while the RSPD mainly reflects population numbers.

A common interpretation of the Early Mesolithic in the region is that the population was small but highly mobile, reflected in large numbers of short lived sites along the coast [e.g. @bjerck2008; @bjerck2017]. This is commonly seen in relation to homogeneous lithic technology over large parts of Scandinavia in this period, and what is believed to have been a rapid colonisation of the Norwegian coastline. High mobility has been seen as necessary both to maintain demographic viability among colonising groups, and for it to be possible to maintain this technological uniformity over large areas [@rowley-conwy2017]. The initial peak in the SSPD could thus be congruent with a process of rapid colonisation of the Norwegian coast, coupled with a high degree of mobility, leading to the overall high site-count in the earliest period. The lacking signal in the RSPD for the first settlement period following the deglaciation could in this view reflect mobility patterns characterised by short occupational duration and corresponding pyrotechnology based on small, short-lived fires [cf. @damm2022] and low population numbers.

Following this earliest phase, @berg-hansen2022 have argued that there is an increase in sites between c. 8000--7000 BCE in the northern part of the study area of this paper. However, their analysis covers a limited area and is based on the aggregation of point-based estimates of shoreline dates, which are aggregated within periods of variable length. Apart from the limitations with using point-based estimates for shoreline dates, the use of disjoint time-intervals of variable duration could cause issues for comparing the relative counts across these units [@bevan2021].

In contrast to the findings of @berg-hansen2022, the analysis undertaken here instead indicates a decline in the number of shoreline dated sites through this time-interval, with possible positive reversals of this trend just after 8000 BCE and around 7000 BCE. It has previously been suggested that a change in mobility patterns occurs from some time around 8300 BCE, involving a reduction in residential mobility [e.g. @damlien2018b; @roalkvam2022]. Furthermore, both genetic data and technological analyses of lithic inventories indicates that the initial phase of human colonisation from the south is followed by an influx, and possible mixing, of people from the east around the same time [@gunther2018; @manninen2021]. If this narrative is coupled with the data observed here, these migratory events could appear to result in a relatively sudden drop in site-count as indicated in the SSPD, which in turn could be related to a change in mobility patterns. The end of this drop also roughly corresponds to the first ^14^C-dates in the RSPD around 8300 BCE -- giving further credence to the suggestion that the appearance of material to be ^14^C-dated is mainly related to cultural and demographic processes as opposed to taphonomic factors.
